Supplementary material

### 2019-02-08 Manuscript/ms.tex
Cleaning up some phrasing

The importance of intrinsic nonlinearities, potentially alongside seasonality and stochastic forcing \cite{Ellner1998, Koelle2004, Grenfell2002}, in governing human disease risk highlight the need to integrate mechanistic modeling into the quantitative tools used to understand the effects of control interventions.
Such mechanistic models can often perform better than complex autoregressive statistical models in describing and forecasting population dynamics \cite{Reilly2005}.
Moreover, in the absence of case-control studies, mechanistic models can provide scenario-based tools that can be used to predict the effect of management actions \cite{Buckland2007}.

Differential equation models provide a natural way to encode the relevant biological mechanisms, although it is important to also account for sources of uncertainty in those models \cite{Hotelling1927, Wikle2010}.
In particular, the values of parameters (e.g., the length of time for which a host is infectious) are often uncertain, which can lead to large uncertainty about the effects of management actions \cite{Elderd2006}.
The structure of the processes themselves can be uncertain \cite{Ellner1998}, and need to be informed by available, often noisy, data.
Bayesian hierarchical modeling provides a coherent framework to account for and integrate this uncertainty across the three levels of the model (data, process, and parameters \cite{Berliner1996, Cressie2009}).

In this paper, we aim to integrate these elements -- a detailed mechanistic model of dengue transmission with a full Bayesian accounting of uncertainty --  to better understand the interplay of forces governing dengue dynamics and their interaction with potential vector control interventions.
We apply this framework to clinical and entomological surveillance data from the city of Vitoria, Brazil.
These data allow us to estimate a latent time series of mosquito mortality rates that modulate the transmission process and link mosquito abundance to human disease.  
We then use the fitted model to explore how perturbations to the mosquito population propagate and interact with the complex epidemiology and nonlinearities involved in dengue transmission.

@@

Explicitly accounting for interactions between serotypes, even only two of the four, leads to a large and complex mechanistic model.
Moreover, because dengue case reports do not include information on serotype, we do not have enough information to inform the dynamics of individual serotypes.
As such, we develop an SEIRS compartment model, similar to \cite{Newton1992, Burattini2008, Pinho2010} which drops serotype-specific dynamics but preserves the period of cross-immunity and the possibility of reinfection.
In this framework, susceptible humans ($S$) become exposed ($E$) through contact with infectious mosquitoes ($V_I$).
Following a latent period ($\frac{1}{\rho}$), exposed humans become infectious ($I$) at which point they can infect susceptible mosquitoes ($V_S$).
Infectious humans recover at rate $\gamma$ and subsequently remain immune ($R$) for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.

Similarly, susceptible mosquitoes ($V_S$) become exposed ($V_E$) by biting infectious humans and pass through a temperature-dependent incubation period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
Because the assumption of an exponentially distributed incubation period (implicit in the specification of a differential equation model) is a poor fit to laboratory observations \cite{Chan2012}, we instead implement a gamma-distributed incubation period by chaining multiple exposed classes together (taking advantage of the fact that a gamma-distributed random variable can be generated through the sum of a collection of exponential random variables with the same rate parameter) \cite{Lloyd2001}.
Total mosquito population size ($V_N$) is controlled by a forced, seasonally varying growth rate ($r(t)$), while the transmission bottleneck is captured with a forced, seasonally varying mortality rate ($d(t)$).
Captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.

@@

We model the centered and log-transformed mosquito mortality rate ($\nu$) and the per-capita mosquito growth rate ($r$) as forced harmonic oscillators with natural periods of one year:
\begin{align}
\frac{d^2\nu}{dt^2} &= -\omega^2 \nu + \epsilon_{\nu t}\\
\frac{d^2 r}{dt^2} &= -\omega^2 r + \epsilon_{rt},
\end{align}
where the angular frequency of the oscillator, $\omega = 2\pi / 52$, the mosquito death rate $d(t) = d_0 \exp(\nu(t))$, and
\begin{align}
\epsilon_{\nu t} & \sim \text{Normal}(0, \sigma^2_{\nu})\\
\epsilon_{rt} & \sim \text{Normal}(0, \sigma^2_r),
\end{align}
for $t = 1 \dots 243$.
These stochastically-forced harmonic oscillators provide a flexible framework for generating smooth seasonal oscillations in the latent mosquito processes \cite{Ramsay2017}.

@@

To connect the differential equation model to the observed case reports, we add an extra state, $C$, that collects the cumulative number of transitions from the exposed to infectious class (assuming that case reporting coincides with the onset of symptoms).
We then model the number of new cases reported in week $t$ ($y_t$) as:
\begin{equation}
y_t  \sim \text{NegBin}(\phi_y (C(t) - C(t-1)), \eta_y),
\end{equation}
where $\phi_y$ is the reporting probability, $C(t) - C(t-1)$ is the number of new infectious humans in week $t$, and $\eta_y$ controls the overdispersion relative to the Poisson distribution.

We similarly model the number of mosquitoes trapped in week $t$ ($q_t$) as:
\begin{equation}
q_t \sim \text{NegBin}(V_{C}(t) - V_{C}(t-1), \eta_q),
\end{equation}
where $V_{C}(t) - V_{C}(t-1)$ is the number of new mosquitoes captured in week $t$, and $\eta_q$ controls overdispersion relative to the Poisson distribution.

@@

Several of the parameters in this model are assumed to be fixed and known (Table 1).
The human population size and average life span (which we use to parameterize the birth/death rate) for Vitoria are taken from the 2010 census.
To maintain identifiability, the transmission rate ($\lambda$) and case reporting probability ($\phi_y$) are also fixed at literature values.
Lastly, the extrinsic incubation period in mosquitoes is modeled as a function of weekly mean temperature and forced with weather station data obtained from WeatherUnderground.

@@

\begin{table}[!ht]
\label{parameters}
\begin{adjustwidth}{-2.25in}{0in} 
\begin{center}
\caption{Model parameters and their values.  Parameters above the rule are fixed, while parameters below the rule are random, with the value giving the prior mean.}
\begin{tabular}{llll}
Parameter & Description & Value & Citation\\
\hline
$N$ & Human population size & 327801 & \cite{vitpop} \\
$1/d$ & Human life-span & 76 years & \cite{vitlong} \\
$\lambda$ & Transmission rate & 4.87 week$^{-1}$ & \cite{Scott2000}\\
$\phi_y$ & Reporting probability & 0.083 & \cite{Silva2016}\\
$1/\rho_{vt}$ & Extrinsic incubation period & $28\exp \left( 0.2 T_t - 8 \right)$ & \cite{Chan2012}\\
$V_{E0}$ & Initial exposed mosquitoes &  0 & \\
$V_{I0}$ & Initial infectious mosquitoes & 0 & \\
\hline
$d_0$ & Baseline mosquito mortality rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
$1/\rho$ & Latent period in host & 0.87 weeks  & \cite{Chan2012}\\
$\gamma$ & Rate of loss of infectiousness & 3.5 week$^{-1}$ & \cite{Nguyet2013}\\
$1/\delta$ & Period of cross-immunity & 97 weeks &  \cite{Reich2013}\\
$\sigma_r$ & Standard deviation of mosquito growth rate forcing & $0^*$ & \\
$\sigma_{\nu}$ & Standard deviation of mosquito mortality rate forcing & $0^*$ & \\
$S_0$ & Proportion initially susceptible & 0.4 & \cite{Cardoso2011a} \\
$E_0$ & Number initially exposed & $100$ & \\
$I_0$ & Number initially infectious & $60$ & \\
$r_0$ & Initial mosquito population growth rate & 0 & \\
$\nu_0$ & Initial unconstrained mosquito mortality rate & 0 & \\
$V_{N0}$ & Initial mosquito population size & $2N$ & \\
$\phi_q$ & Per-trap mosquito capture rate & $2 \times 10^{-6}$ week$^{-1}$ & \\
$\eta_y$ & Overdispersion of case reports & $0^*$ & \\
$\eta_q$ & Overdispersion of mosquito trap counts & $0^*$ & \\
\end{tabular}
\end{center}
\end{adjustwidth}
$^*$ indicates prior mode, rather than mean.
\end{table}

@@

The model captured the observed dynamics of both case reports and mosquito trap counts (Figure \ref{timeseries}).
The estimated posterior median case reports explained 90\% of the variation in the observed time series, while the posterior median mosquito trap counts explained 46\% of the variation in the observed time series. 
In addition, posterior predictive checks showed that the model reproduced the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 6 - 8).
The posterior distributions of the rate of infectious decay ($\gamma$) and the period of cross-immunity ($1/\delta$) did not differ substantially from their priors, suggesting that the Vitoria data contained little additional information about these parameters (Supplemental Figure 3).
The estimated latent period in a human host ($1/\rho$, the expected time it takes for an exposed human to become infectious to biting mosquitoes) is influenced more strongly by the data, with a posterior mean of 1.53 weeks compared to a prior mean of 0.87 weeks.

@@

The effect of a given control intervention seemed to be driven largely by its interaction with, and timing relative to, fluctuations in human immunity.
In particular, mosquito control seemed to be most effective when applied at or near a local minimum in the number of suceptible humans (Figure \ref{timing}).
In addition, human immunity, as controlled by the period of cross-immunity ($1/\delta$), was a major driver of uncertainty in the effect of a given control intervention.
Specifically, we observed a strong negative posterior correlation between the maximum proportion of cases that could be prevented in a given year (i.e., the number of cases prevented by control implemented in the optimal week) and the period of cross-immunity (Figure \ref{immunity}). 
In simulations with a short period of cross-immunity, the susceptible pool replenished quickly, leading to a relatively large pool of susceptibles at any time.
In simulations with a long period of cross-immunity, the susceptible pool replenished more slowly, leaving the population with relatively few susceptibles.
The negative correlation between the effectiveness of control and the period of cross-immunity thus suggests that mosquito control becomes less effective as the overall availability of susceptibles  becomes more limiting.

@@

Infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are often driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity.
In the case of dengue fever, these processes can obscure the relationship between mosquito abundance and human disease.
We found that a mechanistic model with temporary human immunity and seasonal variation in the mosquito mortality rate was able to effectively fit case reports and mosquito surveillance data from Vitoria, Brazil.

The importance of mosquito longevity in driving disease dynamics highlights the potential effectiveness of control efforts that target adult mosquitoes and disrupt transmission by preventing mosquitoes from living long enough to progress through the extrinsic incubation period to the infectious state.
In fact, this forms the basis for much of the theory of adult mosquito control \cite{Burattini2008, Morrison2008, Smith2012}.
Despite this, our simulations suggested that control was likely to be relatively less effective when implemented during an outbreak, due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population \cite{Newton1992, Burattini2008}.
Instead, our results suggest that adult mosquito control efforts are most effective at reducing the disease burden within a given year when applied roughly 30 weeks prior to the start of that year, on the downswing of the previous year's epidemic and seasonal mosquito dynamics.
While the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicate long-term benefits that become apparent in the following outbreak.

The mismatch between the timing of the most effective control and the peak in mosquito abundance suggets that mosquito abundance alone was an insufficient predictor of the effect of mosquito control, at least at the city-scale.
Instead, the effectiveness of mosquito control depended, in large part, on the intrinsic long-term feedbacks and lags driven by the dynamics of temporary human immunity.
In fact, control implemented at a 30 week lag was effective largely due to the way it interacted with the dynamics of the susceptible pool. 
Waiting to implement control until the human susceptible population begins to grow (when the influx of individuals whose previous immunity has waned just outpaces the loss of individuals due to infection) allows for additional immunizing infections to build up as the current outbreak subsides, while also disrupting transmission before the susceptibles that will fuel the next outbreak begin to accumulate.
The fact that effect of mosquito control was mediated by the dynamics of human immunity highlights that dengue, like many other diseases (e.g., cholera \cite{Koelle2004, Koelle2005}), is driven by a complex interplay between seasonal forcing and intrinsic non-linearities.

The Bayesian framework allows for a full accounting of uncertainty across the data, process, and parameter levels of our model \cite{Berliner1996}.
We then carry this uncertainty through to our simulations of control interventions \cite{Elderd2006}.
We found that there can be substantial uncertainty in the proportion of cases prevented by a control intervention (i.e., the width of the ribbons in Figure \ref{control}).
Much of this uncertainty could be attributed to uncertainty in the period of cross-immunity (Figure \ref{immunity}), the posterior distribution of which differed little from our prior based on the estimate of Reich \emph{et al.} \cite{Reich2013}.
Specifically, we found that increasing the period of cross-immunity decreased the proportion of cases prevented by mosquito control.
When the period of cross-immmunity was long (e.g., 200 weeks), the susceptible pool replenished slowly and transmission became limited by the availability of susceptible hosts.
In the face of this limitation, manipulating mosquito abundance had a relatively small effect.
Conversely, when the period of cross-immunity was short (e.g., 50 weeks), susceptibles were abundant and mosquito abundance was instead limiting, in which case reducing mosquito abundance further had a larger effect on transmission.
This interaction between the efficacy of control and human immunity was also demonstrated by ten Bosch \emph{et al.} \cite{TenBosch2016} in their investigation of serotype interactions.
Specifically, they found that adding cross-immunity to models with antibody-dependent enhancement increased the resistance to control (as implemented by reducing the transmission rate).

@@

Given the central role that mosquito lifespan plays in driving dengue dynamics, we modeled fluctuations in the mosquito mortality rate using a flexible, forced harmonic oscillator.
This encodes our prior expectation that the mosquito mortality rate should be driven by seasonal environmental forcing (i.e., it should oscillate with a period of one year), while allowing the flexibility to diverge from the simple oscillator and learn from the case reports data.
The fact that the $\epsilon_{\nu}$ forcing terms are the only source of variability in the transmission process, however, means that the estimated mosquito mortality time series could be soaking up other sources of stochasticity or model misspecification.
Hooker and Ellner \cite{Hooker2015} provide a framework for diagnosing such model misspecification in differential equation models using forcing functions similar to our implementation of the $\epsilon_{\nu}$.
In that framework, Hooker and Ellner \cite{Hooker2015} estimate nonparametric forcing functions that modify a fitted differential equation model to provide a good fit to the data.
These forcing functions serve as residuals on the time derivatives, and can be more readily interpreted as indicators of lack-of-fit than residuals on the state variables \cite{Hotelling1927, Hooker2015}.
We do not employ the same explicit goodness-of-fit testing framework as \cite{Hooker2015}, but we can inspect our estimated $\epsilon_{\nu}$ forcing terms in the same spirit.

The periodic structure in the posterior means of the $\epsilon_{\nu t}$ suggests that these terms are accounting for more than just noise, and there may be some unmodeled process influencing fluctuations in mosquito mortality and/or transmission.
Following Hooker and Ellner, we can explore whether this process is likely to result from misspecification of the rates of change of the existing state variables (indicated by a dependence of $\epsilon_{\nu t}$ on other state variables), or from missing state variables altogether (indicated by an additional dependence of $\epsilon_{\nu t}$ on its own lagged values). 
The lack of any apparent relationship between the forcing terms and any of the estimated state variables, combined with the dependence of $\epsilon_{\nu t}$ on previous values (as apparent through the periodic structure), suggest that unmodeled state variables may be the more likely driver of model misspecification.
These unmodeled components could include additional mosquito population dynamic processes (e.g., aquatic stage dynamics, environmental drivers, or control interventions), or epidiemiological processes (e.g., multiple circulating serotypes of the dengue virus).

Despite this suggestion of potential model misspecification, our estimated mosquito mortalities nonetheless fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
Moreover, the fact that our estimated mortality rate increased with temperature also broadly agrees with the empirical literature on mosquito survival \cite{Yang2009, Brady2013}.
This suggests that regardless of unexplained structure in the forcing terms, the pattern of case reports is still very well described by realistic fluctuations in the mosquito mortality rate.

@@

We view our simulations of this additional pulse of control primarily as a tool to explore and exploit the long-term feedbacks in the city's epidemiological system, whereas existing control is deployed largely to reduce local (in space and time) disease risk.
The frequent use of spatially-targeted mosquito control highlights the potential for spatial heterogeneities in disease risk within a city.
In particular, structured human movement within the city is likely to induce heterogeneous human-mosquito mixing \cite{Adams2009, Cosner2009a, Stoddard2009}.
In addition, spatial variability in socioeconomic factors within the city may also modulate the extent to which mosquitoes in different parts of the city contribute to disease spread \cite{Mondini2008, Honorio2009, Hu2012, DeMattosAlmeida2007}.
Although our model does not account for these heterogeneities, it is nontheless able to capture the city-wide dynamics reasonably well, suggesting that there may be sufficient mixing to appear homogeneous at the city scale.
Further, although our mechanistic model may be able to suggest when a city-wide intervention is likely to be effective, to make the best use of limited resources, spatial prioritization may still be necessary.
Our results emphasize that work to develop the tools necessary for this prioritization should focus not just on prioritization based on the distribution of mosquito abundance, but also on the distribution of human immunity.

>>>>

The importance of intrinsic nonlinearities, potentially alongside seasonality and stochastic forcing \cite{Ellner1998, Koelle2004, Grenfell2002}, in governing human disease risk highlights the need to integrate mechanistic modeling into the quantitative tools used to understand the effects of control interventions.
Such mechanistic models can often perform better than complex autoregressive statistical models in describing and forecasting population dynamics \cite{Reilly2005}.
Moreover, in the absence of case-control studies, mechanistic models can provide scenario-based tools that can be used to predict the effect of management actions \cite{Buckland2007}.

Differential equation models provide a natural way to encode the relevant biological mechanisms, although it is important to also account for sources of uncertainty in those models \cite{Hotelling1927, Wikle2010}.
In particular, the values of parameters (e.g., the length of time for which a host is infectious) are often uncertain, which can lead to large uncertainty about the effects of management actions \cite{Elderd2006}.
The structure of the processes themselves can be uncertain \cite{Ellner1998}, and needs to be informed by available, often noisy, data.
Bayesian hierarchical modeling provides a coherent framework to account for and integrate this uncertainty across the three levels of the model (data, process, and parameters \cite{Berliner1996, Cressie2009}).

In this paper, we aim to integrate these elements -- a detailed mechanistic model of dengue transmission with a full Bayesian accounting of uncertainty --  to better understand the interplay of forces governing dengue dynamics and their interaction with potential vector control interventions.
We apply this framework to clinical and entomological surveillance data from the city of Vitoria, Brazil.
These data allow us to estimate a latent time series of mosquito mortality rates that modulate the transmission process and link mosquito abundance to human disease.  
We then use the fitted model to explore how perturbations to the mosquito population propagate and interact with the nonlinearities of dengue transmission in order to better inform mosquito control efforts.

@@

Explicitly accounting for interactions between serotypes, even only two of the four, leads to a large and complex mechanistic model.
Moreover, because dengue case reports do not identify serotype, there is not enough information in the data to inform the dynamics of individual serotypes.
As such, we developed an SEIRS compartment model, similar to \cite{Newton1992, Burattini2008, Pinho2010} which drops serotype-specific dynamics but preserves the period of cross-immunity and the possibility of reinfection.
In this framework, susceptible humans ($S$) become exposed ($E$) through contact with infectious mosquitoes ($V_I$).
Following a latent period ($\frac{1}{\rho}$), exposed humans become infectious ($I$) at which point they can infect susceptible mosquitoes ($V_S$).
Infectious humans recover at rate $\gamma$ and subsequently remain immune ($R$) for a period ($\frac{1}{\delta}$) after which they re-enter the susceptible class.

Similarly, susceptible mosquitoes ($V_S$) become exposed ($V_E$) by biting infectious humans and pass through a temperature-dependent incubation period ($\frac{1}{\rho_{vt}}$) before becoming infectious ($V_I$).
Because the assumption of an exponentially distributed incubation period (implicit in the specification of a differential equation model) is a poor fit to laboratory observations \cite{Chan2012}, we instead implemented a gamma-distributed incubation period by chaining multiple exposed classes together (taking advantage of the fact that a gamma-distributed random variable can be generated through the sum of a collection of exponential random variables with the same rate parameter) \cite{Lloyd2001}.
Total mosquito population size ($V_N$) is controlled by a forced, seasonally varying growth rate ($r(t)$), while the transmission bottleneck is captured with a forced, seasonally varying mortality rate ($d(t)$).
Captured mosquitoes ($V_C$) accumulate at rate $\phi_q \tau_t$, where $\phi_q$ is the per-trap capture rate, and $\tau_t$ is the number of traps deployed in week $t$.

@@

We modeled the centered and log-transformed mosquito mortality rate ($\nu$) and the per-capita mosquito growth rate ($r$) as forced harmonic oscillators with natural periods of one year:
\begin{align}
\frac{d^2\nu}{dt^2} &= -\omega^2 \nu + \epsilon_{\nu t}\\
\frac{d^2 r}{dt^2} &= -\omega^2 r + \epsilon_{rt},
\end{align}
where the angular frequency of the oscillator, $\omega = 2\pi / 52$, the mosquito death rate $d(t) = d_0 \exp(\nu(t))$, and
\begin{align}
\epsilon_{\nu t} & \sim \text{Normal}(0, \sigma^2_{\nu})\\
\epsilon_{rt} & \sim \text{Normal}(0, \sigma^2_r),
\end{align}
for $t = 1 \dots 243$.
These stochastically-forced harmonic oscillators provide a flexible framework for generating smooth seasonal oscillations in the latent mosquito processes \cite{Ramsay2017}.

@@

To connect the differential equation model to the observed case reports, we added an extra state, $C$, that collects the cumulative number of transitions from the exposed to infectious class (assuming that case reporting coincides with the onset of symptoms).
We then modeled the number of new cases reported in week $t$ ($y_t$) as:
\begin{equation}
y_t  \sim \text{NegBin}(\phi_y (C(t) - C(t-1)), \eta_y),
\end{equation}
where $\phi_y$ is the reporting probability, $C(t) - C(t-1)$ is the number of new infectious humans in week $t$, and $\eta_y$ controls the overdispersion relative to the Poisson distribution.

We similarly modeled the number of mosquitoes trapped in week $t$ ($q_t$) as:
\begin{equation}
q_t \sim \text{NegBin}(V_{C}(t) - V_{C}(t-1), \eta_q),
\end{equation}
where $V_{C}(t) - V_{C}(t-1)$ is the number of new mosquitoes captured in week $t$, and $\eta_q$ controls overdispersion relative to the Poisson distribution.

@@

Several of the parameters in this model are assumed to be fixed and known (Table 1).
The human population size and average life span (which we use to parameterize the birth/death rate) for Vitoria were taken from the 2010 census.
To maintain identifiability, the transmission rate ($\lambda$) and case reporting probability ($\phi_y$) were also fixed at literature values.
Lastly, the extrinsic incubation period in mosquitoes was modeled as a function of weekly mean temperature and forced with weather station data obtained from WeatherUnderground.

@@

\begin{table}[!ht]
\label{parameters}
\begin{adjustwidth}{-2.25in}{0in} 
\begin{center}
\caption{Model parameters and their values.  Parameters above the rule are fixed, while parameters below the rule are random, with the value giving the prior mean. See Supplemental Material for a full description of all priors.}
\begin{tabular}{llll}
Parameter & Description & Value & Citation\\
\hline
$N$ & Human population size & 327801 & \cite{vitpop} \\
$1/d$ & Human life-span & 76 years & \cite{vitlong} \\
$\lambda$ & Transmission rate & 4.87 week$^{-1}$ & \cite{Scott2000}\\
$\phi_y$ & Reporting probability & 0.083 & \cite{Silva2016}\\
$1/\rho_{vt}$ & Extrinsic incubation period & $28\exp \left( 0.2 T_t - 8 \right)$ & \cite{Chan2012}\\
$V_{E0}$ & Initial exposed mosquitoes &  0 & \\
$V_{I0}$ & Initial infectious mosquitoes & 0 & \\
\hline
$d_0$ & Baseline mosquito mortality rate & 1.47 week$^{-1}$ & \cite{Brady2013} \\
$1/\rho$ & Latent period in host & 0.87 weeks  & \cite{Chan2012}\\
$\gamma$ & Rate of loss of infectiousness & 3.5 week$^{-1}$ & \cite{Nguyet2013}\\
$1/\delta$ & Period of cross-immunity & 97 weeks &  \cite{Reich2013}\\
$\sigma_r$ & Standard deviation of mosquito growth rate forcing & $0^*$ & \\
$\sigma_{\nu}$ & Standard deviation of mosquito mortality rate forcing & $0^*$ & \\
$S_0$ & Proportion initially susceptible & 0.4 & \cite{Cardoso2011a} \\
$E_0$ & Number initially exposed & $100$ & \\
$I_0$ & Number initially infectious & $60$ & \\
$r_0$ & Initial mosquito population growth rate & 0 & \\
$\nu_0$ & Initial unconstrained mosquito mortality rate & 0 & \\
$V_{N0}$ & Initial mosquito population size & $2N$ & \\
$\phi_q$ & Per-trap mosquito capture rate & $2 \times 10^{-6}$ week$^{-1}$ & \\
$\eta_y$ & Overdispersion of case reports & $0^*$ & \\
$\eta_q$ & Overdispersion of mosquito trap counts & $0^*$ & \\
\end{tabular}
\end{center}
\end{adjustwidth}
$^*$ indicates prior mode, rather than mean.
\end{table}

@@

The model captured the observed dynamics of both case reports and mosquito trap counts (Figure \ref{timeseries}).
The estimated posterior median case reports explained 90\% of the variation in the observed time series, while the posterior median mosquito trap counts explained 46\% of the variation in the observed time series. 
In addition, posterior predictive checks showed that the model reproduced the total number of cases reported and mosquitoes captured as well as the autocorrelation structure of both time series (with the exception of slightly underestimating the autocorrelation for short lags, Supplemental Figures 6 - 8).
The posterior distributions of the rate of infectious decay ($\gamma$) and the period of cross-immunity ($1/\delta$) did not differ substantially from their priors, suggesting that the Vitoria data contained little additional information about these parameters (Supplemental Figure 3).
The estimated latent period in a human host ($1/\rho$, the expected time it takes for an exposed human to become infectious to biting mosquitoes) was influenced more strongly by the data, with a posterior mean of 1.53 weeks compared to a prior mean of 0.87 weeks.

@@

The effect of a given control intervention seemed to be driven largely by its interaction with, and timing relative to, fluctuations in human immunity.
In particular, mosquito control seemed to be most effective when applied at or near a local minimum in the number of suceptible humans (Figure \ref{timing}).
In addition, human immunity, as controlled by the period of cross-immunity ($1/\delta$), was a major driver of uncertainty in the effect of a given control intervention.
Specifically, we observed a strong negative posterior correlation between the maximum proportion of cases that could be prevented in a given year (i.e., the number of cases prevented by control implemented in the optimal week) and the period of cross-immunity (Figure \ref{immunity}). 
In simulations with a short period of cross-immunity, the susceptible pool replenished quickly, leading to a relatively large pool of susceptibles at any time.
In simulations with a long period of cross-immunity, the susceptible pool replenished more slowly, leaving the population with relatively few susceptibles.
The negative correlation between the effectiveness of control and the period of cross-immunity thus suggests that mosquito control was less effective when the overall availability of susceptibles was more limiting.

@@

The dynamics of dengue fever, like those of many infectious diseases \cite{Ellner1998,Koelle2004} and ecological systems \cite{Bjornstad2001}, are driven by the combined efforts of intrinsic non-linearities, seasonality, and stochasticity.
In the dengue system, these drivers include waning human immunity, the race between the extrinsic incubation period of the virus and mosquito lifespan, and the seasonal and stochastic variation in these rates.
These processes are particularly important for establishing the relationship between mosquito abundance -- and efforts to control that abundance -- and human disease.

The importance of mosquito longevity in driving disease dynamics highlights the potential effectiveness of control efforts that target adult mosquitoes and disrupt transmission by preventing mosquitoes from living long enough to progress through the extrinsic incubation period to the infectious state.
In fact, this forms the basis for much of the theory of adult mosquito control \cite{Burattini2008, Morrison2008, Smith2012}.
Despite this, our simulations suggest that control was relatively less effective when implemented during an outbreak, likely due to the large number of infectious human hosts available to transmit to the remaining (and rapidly rebounding) mosquito population \cite{Newton1992, Burattini2008}.
Instead, we found that adult mosquito control efforts were most effective at reducing the disease burden within a given year when applied roughly 30 weeks prior to the start of that year, on the downswing of the previous year's epidemic and seasonal mosquito dynamics.
While the immediate effects of mosquito control implemented during the decline phase of an outbreak may be masked by the natually fading transmission intensity \cite{Stoddard2014}, our results nonetheless indicated long-term benefits that carried through to the following outbreak.

The mismatch between the timing of the most effective control and the peak in mosquito abundance suggets that mosquito abundance alone was an insufficient predictor of the effect of mosquito control, at least at the city-scale.
Instead, the effectiveness of mosquito control depended, in large part, on the intrinsic long-term feedbacks and lags driven by the dynamics of temporary human immunity.
In fact, control implemented at a 30 week lag was effective largely due to the way it interacted with the dynamics of the susceptible pool. 
Implementing control just as the human susceptible population begins to grow (when the influx of individuals whose previous immunity has waned just outpaces the loss of individuals due to infection) allows for additional immunizing infections to build up as the current outbreak subsides, while also disrupting transmission before the susceptibles that will fuel the next outbreak begin to accumulate.
The fact that effect of mosquito control was mediated by the dynamics of human immunity highlights that dengue, like many other diseases (e.g., cholera \cite{Koelle2004, Koelle2005}), is driven by a complex interplay between seasonal forcing and intrinsic non-linearities.

The Bayesian framework allowed for a full accounting of uncertainty across the data, process, and parameter levels of our model \cite{Berliner1996}.
We then carried this uncertainty through to our simulations of control interventions \cite{Elderd2006}.
We found that there can be substantial uncertainty in the proportion of cases prevented by a control intervention (i.e., the width of the ribbons in Figure \ref{control}).
Much of this uncertainty could be attributed to uncertainty in the period of cross-immunity (Figure \ref{immunity}), the posterior distribution of which differed little from our prior based on the estimate of Reich \emph{et al.} \cite{Reich2013}.
Specifically, we found that increasing the period of cross-immunity decreased the proportion of cases prevented by mosquito control.
When the period of cross-immmunity was long (e.g., 200 weeks), the susceptible pool replenished slowly and transmission became limited by the availability of susceptible hosts.
In the face of this limitation, manipulating mosquito abundance had a relatively small effect.
Conversely, when the period of cross-immunity was short (e.g., 50 weeks), susceptibles were abundant and mosquito abundance was instead limiting, in which case reducing mosquito abundance further had a larger effect on transmission.
This interaction between the efficacy of control and human immunity was also demonstrated by ten Bosch \emph{et al.} \cite{TenBosch2016} in their investigation of serotype interactions.
Specifically, they found that adding cross-immunity to models with antibody-dependent enhancement increased the resistance to control (as implemented by reducing the transmission rate).

@@

Given the central role that mosquito lifespan plays in driving dengue dynamics, we modeled fluctuations in the mosquito mortality rate using a flexible, forced harmonic oscillator.
This encoded our prior expectation that the mosquito mortality rate should be driven by seasonal environmental forcing (i.e., it should oscillate with a period of one year), while allowing the flexibility to diverge from the simple oscillator and learn from the case reports data.
The fact that the $\epsilon_{\nu}$ forcing terms were the only source of variability in the transmission process, however, means that the estimated mosquito mortality time series could have been soaking up other sources of stochasticity or model misspecification.
Hooker and Ellner \cite{Hooker2015} provide a framework for diagnosing such model misspecification in differential equation models using forcing functions similar to our implementation of the $\epsilon_{\nu}$.
In that framework, Hooker and Ellner \cite{Hooker2015} estimate nonparametric forcing functions that modify a fitted differential equation model to provide a good fit to the data.
These forcing functions serve as residuals on the time derivatives, and can be more readily interpreted as indicators of lack-of-fit than residuals on the state variables \cite{Hotelling1927, Hooker2015}.
We do not employ the same explicit goodness-of-fit testing framework as \cite{Hooker2015}, but we can inspect our estimated $\epsilon_{\nu}$ forcing terms in the same spirit.

The periodic structure in the posterior means of the $\epsilon_{\nu t}$ suggests that these terms were accounting for more than just noise, and there may have been some unmodeled process influencing fluctuations in mosquito mortality and/or transmission.
Following Hooker and Ellner, we can explore whether this process is likely to result from misspecification of the rates of change of the existing state variables (indicated by a dependence of $\epsilon_{\nu t}$ on other state variables), or from missing state variables altogether (indicated by an additional dependence of $\epsilon_{\nu t}$ on its own lagged values). 
The lack of any apparent relationship between the forcing terms and any of the estimated state variables, combined with the dependence of $\epsilon_{\nu t}$ on previous values (as apparent through the periodic structure), suggest that unmodeled state variables may be the more likely driver of model misspecification.
These unmodeled components could include additional mosquito population dynamic processes (e.g., aquatic stage dynamics, environmental drivers, or control interventions), or epidiemiological processes (e.g., multiple circulating serotypes of the dengue virus).

Despite this suggestion of potential model misspecification, our estimated mosquito mortalities nonetheless fell within the reasonable range from the literature \cite{Maciel-de-Freitas2008, Brady2013}.
Moreover, the fact that our estimated mortality rate increased with temperature also broadly agrees with the empirical literature on mosquito survival \cite{Yang2009, Brady2013}.
This suggests that regardless of unexplained structure in the forcing terms, the pattern of case reports was still very well described by realistic fluctuations in the mosquito mortality rate.

@@

We view our simulations of this additional pulse of control primarily as a tool to explore and exploit the long-term feedbacks in the city's epidemiological system, whereas existing control is deployed largely to reduce local (in space and time) disease risk.
The frequent use of spatially-targeted mosquito control highlights the potential for spatial heterogeneities in disease risk within a city.
In particular, structured human movement within the city is likely to induce heterogeneous human-mosquito mixing \cite{Adams2009, Cosner2009a, Stoddard2009}.
In addition, spatial variability in socioeconomic factors within the city may also modulate the extent to which mosquitoes in different parts of the city contribute to disease spread \cite{Mondini2008, Honorio2009, Hu2012, DeMattosAlmeida2007}.
Although our model does not account for these heterogeneities, it was nontheless able to capture the city-wide dynamics reasonably well, suggesting that there may be sufficient mixing to appear homogeneous at the city scale.
Further, although our mechanistic model may be able to suggest when a city-wide intervention is likely to be effective, to make the best use of limited resources, spatial prioritization may still be necessary.
Our results emphasize that work to develop the tools necessary for this prioritization should focus not just on prioritization based on the distribution of mosquito abundance, but also on the distribution of human immunity.
